Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-18 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

%---------------------------------------------------------------------------------------------------%
%% INTRODUCTION
\section{Introduction and Background}
Naturally occurring porous and granular materials, such as soil, sand, and clay, play a pivotal role in regulating earth's water resources by filtering contaminants and, over long timescales, supplying fresh water.  Not only is this process essential for human water resources, but also for the ecology of rivers, estuaries, and other natural habitats.  Natural water resources, however, have been placed under enormous pressure by human population growth and associated activities that can compromise the natural filtration cycle, for example expansion of industry, urbanization, pollution, and climate change. See the UN report on water resources for further detail on the threats to water resources~\cite{UNwater}.
% Could also include USGS if needed.

An understanding of these filtration and contamination processes relies on the complex phenomenon of transport in porous media. Further, the action of continually flowing groundwater can alter medium properties through erosive and transportive processes. Thus, it becomes necessary to understand transport through complex media whose properties change {\em dynamically} in response to the intervening fluid flow. These effects are most noticeable during rapid events, like the gravitational collapse of a sinkhole~\cite{sandhu2018fate}, during which medium properties can change dramatically over short timescales. Though less obvious, the accumulation of slower process, such as mechanical or chemical erosion, can also substantially alter medium properties.

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow can alter porous-media properties and to characterize the associated changes in transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

In particular, we view the accumulated effects of slow processes like erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}.

>>>>

%---------------------------------------------------------------------------------------------------%
%% INTRODUCTION
\section{Introduction and Background}
Naturally occurring porous and granular materials, such as soil, sand, and clay, play a pivotal role in regulating earth's water resources by filtering contaminants and, over long timescales, supplying fresh water.  Not only is this process essential for human water resources, but also for the ecology of rivers, estuaries, and other natural habitats.  Natural water resources, however, have been placed under enormous pressure by human population growth and associated activities that can compromise the natural filtration cycle, for example expansion of industry, urbanization, pollution, and climate change. See the UN report on water resources for further detail on these looming threats~\cite{UNwater}.
% Could also include USGS if needed.

An understanding of these filtration and contamination processes relies on the complex phenomenon of transport and dispersion in porous media. Further, the action of continually flowing groundwater can alter medium properties through erosive, transportive, and other processes. Thus, it becomes necessary to understand transport through complex media whose properties change {\em dynamically} in response to the intervening fluid flow. These effects are most noticeable during rapid events, like the gravitational collapse of a sinkhole~\cite{sandhu2018fate}, during which medium properties can change dramatically over short timescales. Though less obvious, the accumulation of slower process, such as mechanical or chemical erosion, can also substantially alter medium properties.

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow alters porous-media properties and to characterize the associated changes in dispersive transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Throughout this proposal, we also refer to chemically-driven erosion as dissolution. Some examples in the figure include the rough surface of limestone pavement, a sinkhole, an erosion-curved conduit, and finally an image of the small-scale structure of soil, exhibiting polydispersity and anisotropy, resulting from a combination of erosive, compaction, and transport properties
